---
title: "Respiratory Surveillance · England"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: cosmo
---

```{r setup, include=FALSE}

# 1. PACKAGES

library(flexdashboard)
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(ggplot2)
library(plotly)
library(forecast)
library(DT)
library(scales)
library(htmltools)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

```


<style>

/* =========================================================
   GLOBAL
   ========================================================= */

body {
  background: #F6F4EF;
  color: #1C2522;
  font-family:
    -apple-system,
    BlinkMacSystemFont,
    "Segoe UI",
    Helvetica,
    Arial,
    sans-serif;
}

.container-fluid {
  background: #F6F4EF;
}

.chart-wrapper {
  background: transparent;
  border: none;
  box-shadow: none;
}

.chart-stage {
  overflow: visible;
}

.dashboard-row {
  margin-bottom: 16px;
}

p {
  color: #53605B;
  line-height: 1.7;
  font-size: 15px;
}


/* =========================================================
   NAVIGATION
   ========================================================= */

.navbar-inverse {
  background: #F6F4EF;
  border-color: #D8D5CD;
}

.navbar-inverse .navbar-brand {
  color: #17201D !important;
  font-weight: 650;
  letter-spacing: -0.02em;
}

.navbar-inverse .navbar-nav > li > a {
  color: #68716D !important;
  font-size: 13px;
  font-weight: 600;
  letter-spacing: 0.02em;
}

.navbar-inverse .navbar-nav > li > a:hover {
  color: #17201D !important;
  background: transparent !important;
}

.navbar-inverse .navbar-nav > .active > a,
.navbar-inverse .navbar-nav > .active > a:hover {
  color: #17201D !important;
  background: transparent !important;
  border-bottom: 2px solid #17201D;
}


/* =========================================================
   HERO
   ========================================================= */

.hero {
  padding: 46px 28px 30px 28px;
  max-width: 1100px;
}

.eyebrow {
  color: #66736E;
  font-size: 11px;
  font-weight: 750;
  letter-spacing: 0.16em;
  text-transform: uppercase;
  margin-bottom: 14px;
}

.hero h1 {
  color: #17201D;
  font-size: 44px;
  line-height: 1.05;
  font-weight: 650;
  letter-spacing: -0.045em;
  margin-top: 0;
  margin-bottom: 18px;
}

.hero-copy {
  color: #58645F;
  font-size: 17px;
  line-height: 1.6;
  max-width: 760px;
  margin-bottom: 22px;
}

.update-line {
  display: inline-block;
  color: #53605B;
  font-size: 12px;
  font-weight: 600;
  border-top: 1px solid #D8D5CD;
  padding-top: 12px;
}


/* =========================================================
   SECTION INTRO
   ========================================================= */

.section-intro {
  padding: 12px 12px 4px 12px;
}

.section-label {
  color: #69746F;
  font-size: 10px;
  font-weight: 750;
  letter-spacing: 0.16em;
  text-transform: uppercase;
  margin-bottom: 8px;
}

.section-intro h2 {
  color: #17201D;
  font-size: 30px;
  font-weight: 650;
  letter-spacing: -0.035em;
  margin-top: 0;
  margin-bottom: 10px;
}

.section-intro p {
  max-width: 760px;
  margin-bottom: 10px;
}


/* =========================================================
   METRIC CARDS
   ========================================================= */

.metric-grid {
  display: grid;
  grid-template-columns: repeat(4, minmax(0, 1fr));
  gap: 14px;
  padding: 4px 12px 20px 12px;
}

.metric-card {
  background: #FFFFFF;
  border: 1px solid #DDD9D1;
  border-radius: 12px;
  padding: 19px 20px 17px 20px;
  min-height: 112px;
}

.metric-label {
  color: #707A76;
  font-size: 10px;
  font-weight: 750;
  letter-spacing: 0.12em;
  text-transform: uppercase;
}

.metric-value {
  color: #17201D;
  font-size: 30px;
  line-height: 1.1;
  font-weight: 650;
  letter-spacing: -0.04em;
  margin-top: 9px;
}

.metric-note {
  color: #7A837F;
  font-size: 11px;
  margin-top: 7px;
}


/* =========================================================
   PATHOGEN CARDS
   ========================================================= */

.pathogen-grid {
  display: grid;
  grid-template-columns: repeat(3, minmax(0, 1fr));
  gap: 14px;
  padding: 4px 12px 20px 12px;
}

.pathogen-card {
  background: #FFFFFF;
  border: 1px solid #DDD9D1;
  border-radius: 12px;
  padding: 19px 20px;
  min-height: 150px;
}

.pathogen-name {
  color: #4E5A55;
  font-size: 11px;
  font-weight: 750;
  letter-spacing: 0.11em;
  text-transform: uppercase;
}

.pathogen-value {
  color: #17201D;
  font-size: 33px;
  font-weight: 650;
  letter-spacing: -0.04em;
  margin: 12px 0 15px 0;
}

.pathogen-unit {
  color: #69746F;
  font-size: 12px;
  font-weight: 500;
}

.status {
  display: inline-block;
  border-radius: 999px;
  padding: 6px 9px;
  font-size: 10px;
  font-weight: 750;
  letter-spacing: 0.07em;
  text-transform: uppercase;
}

.status-normal {
  color: #326657;
  background: #E8F1ED;
}

.status-signal {
  color: #8F352D;
  background: #F7E8E4;
}

.status-unknown {
  color: #666B69;
  background: #EEEDE9;
}


/* =========================================================
   INTERPRETATION BOXES
   ========================================================= */

.note-box {
  background: #ECEAE4;
  border-left: 3px solid #687A72;
  border-radius: 4px;
  padding: 17px 20px;
  margin: 10px 12px 22px 12px;
  color: #47534E;
  font-size: 14px;
  line-height: 1.65;
}

.note-box strong {
  color: #17201D;
}

.signal-box {
  background: #F6EAE7;
  border-left: 3px solid #A94A40;
}


/* =========================================================
   PLOTS
   ========================================================= */

.chart-title {
  color: #17201D;
  font-size: 14px;
  font-weight: 650;
  border-bottom: none;
}

svg text {
  font-family:
    -apple-system,
    BlinkMacSystemFont,
    "Segoe UI",
    Helvetica,
    Arial,
    sans-serif !important;
}


/* =========================================================
   TABLES
   ========================================================= */

table.dataTable {
  background: #FFFFFF;
  border-collapse: collapse !important;
}

table.dataTable thead th {
  color: #53605B;
  font-size: 10px;
  text-transform: uppercase;
  letter-spacing: 0.08em;
  border-bottom: 1px solid #D8D5CD !important;
}

table.dataTable tbody td {
  color: #26312D;
  border-top: 1px solid #EEECE7 !important;
}


/* =========================================================
   METHOD CARDS
   ========================================================= */

.method-card {
  background: #FFFFFF;
  border: 1px solid #DDD9D1;
  border-radius: 12px;
  padding: 23px 24px;
  margin: 0 12px 16px 12px;
}

.method-card h3 {
  color: #17201D;
  font-size: 19px;
  font-weight: 650;
  margin-top: 3px;
}

.method-card p,
.method-card li {
  color: #56625D;
  line-height: 1.65;
}


/* =========================================================
   RESPONSIVE
   ========================================================= */

@media (max-width: 900px) {

  .metric-grid {
    grid-template-columns: repeat(2, 1fr);
  }

  .pathogen-grid {
    grid-template-columns: repeat(2, 1fr);
  }

  .hero h1 {
    font-size: 36px;
  }
}

@media (max-width: 600px) {

  .metric-grid,
  .pathogen-grid {
    grid-template-columns: 1fr;
  }

  .hero {
    padding-left: 18px;
    padding-right: 18px;
  }
}

</style>


```{r data, include=FALSE}

# 2. LOAD RESPIRATORY DATA

respiratory <- read_csv(
  "data/processed/respiratory_surveillance_long.csv",
  show_col_types = FALSE
) |>
  mutate(
    date = as.Date(date)
  )


# 3. PATHOGEN COLOURS

pathogen_palette <- c(
  "Influenza" = "#B44A42",
  "RSV" = "#C9853E",
  "Rhinovirus" = "#327565",
  "hMPV" = "#7066A2",
  "Adenovirus" = "#52789A",
  "Parainfluenza" = "#9A617D"
)


pathogen_order <- c(
  "Influenza",
  "RSV",
  "Rhinovirus",
  "hMPV",
  "Adenovirus",
  "Parainfluenza"
)


# 4. RESPIRATORY SEASON

respiratory_signals <- respiratory |>
  mutate(
    season_start_year =
      if_else(
        month(date) >= 7,
        year(date),
        year(date) - 1
      ),

    season =
      paste0(
        season_start_year,
        "/",
        substr(
          season_start_year + 1,
          3,
          4
        )
      ),

    season_start =
      as.Date(
        paste0(
          season_start_year,
          "-07-01"
        )
      ),

    week_of_season =
      floor(
        as.numeric(
          date - season_start
        ) / 7
      ) + 1
  )


# 5. CURRENT SEASON

current_season <- respiratory_signals |>
  filter(
    date == max(date)
  ) |>
  distinct(
    season
  ) |>
  pull(
    season
  )


# 6. HISTORICAL BASELINE

historical_baseline <- respiratory_signals |>
  filter(
    season != current_season
  ) |>
  group_by(
    disease,
    week_of_season
  ) |>
  summarise(
    expected =
      median(
        positivity,
        na.rm = TRUE
      ),

    upper_threshold =
      quantile(
        positivity,
        probs = 0.95,
        na.rm = TRUE
      ),

    .groups = "drop"
  )


# 7. SIGNALS

signals_all <- respiratory_signals |>
  left_join(
    historical_baseline,
    by = c(
      "disease",
      "week_of_season"
    )
  ) |>
  mutate(
    signal =
      positivity > upper_threshold
  )


signals <- signals_all |>
  group_by(
    disease
  ) |>
  slice_max(
    date,
    n = 1,
    with_ties = FALSE
  ) |>
  ungroup()


# 8. LATEST DATA

latest_date <- max(
  respiratory$date,
  na.rm = TRUE
)


latest_data <- respiratory |>
  filter(
    date == latest_date
  )


recent_data <- respiratory |>
  filter(
    date >= latest_date - weeks(52)
  )


signal_recent <- signals_all |>
  filter(
    date >= latest_date - weeks(26)
  )


# 9. CURRENT SIGNAL SUMMARY

current_signal_count <- signals |>
  filter(
    signal %in% TRUE
  ) |>
  nrow()


current_signal_names <- signals |>
  filter(
    signal %in% TRUE
  ) |>
  pull(
    disease
  )


# 10. NHS 111 DATA

syndromic <- read_csv(
  "data/processed/syndromic_influenza_aligned.csv",
  show_col_types = FALSE
) |>
  mutate(
    week_start = as.Date(week_start)
  )


latest_nhs111 <- syndromic |>
  slice_max(
    week_start,
    n = 1,
    with_ties = FALSE
  )


nhs111_recent <- syndromic |>
  arrange(
    week_start
  ) |>
  mutate(
    weekly_change =
      weekly_calls -
      lag(weekly_calls),

    percent_change =
      100 *
      weekly_change /
      lag(weekly_calls)
  )


latest_nhs111_change <- nhs111_recent |>
  slice_max(
    week_start,
    n = 1,
    with_ties = FALSE
  )


# 11. CURRENT FORECAST FUNCTION

forecast_current <- function(data) {

  data <- data |>
    arrange(
      date
    )

  disease_name <-
    unique(
      data$disease
    )

  y <- ts(
    data$positivity,
    frequency = 52
  )

  model <- auto.arima(
    y,
    seasonal = TRUE,
    stepwise = TRUE,
    approximation = TRUE
  )

  fc <- forecast(
    model,
    h = 4,
    level = c(
      80,
      95
    )
  )

  last_date <-
    max(
      data$date
    )

  forecast_dates <- seq(
    last_date + weeks(1),
    by = "1 week",
    length.out = 4
  )

  tibble(
    disease =
      disease_name,

    date =
      forecast_dates,

    horizon =
      1:4,

    forecast =
      as.numeric(
        fc$mean
      ),

    lower_80 =
      as.numeric(
        fc$lower[, "80%"]
      ),

    upper_80 =
      as.numeric(
        fc$upper[, "80%"]
      ),

    lower_95 =
      as.numeric(
        fc$lower[, "95%"]
      ),

    upper_95 =
      as.numeric(
        fc$upper[, "95%"]
      )
  )
}


# 12. CURRENT FORECASTS

current_forecasts <- respiratory |>
  group_by(
    disease
  ) |>
  group_split() |>
  lapply(
    forecast_current
  ) |>
  bind_rows()


forecast_history <- respiratory |>
  filter(
    date >= latest_date - weeks(26)
  )


# 13. LEAD-LAG DATA

lead_lag <- read_csv(
  "outputs/tables/change_lag_correlations.csv",
  show_col_types = FALSE
)


# 14. COMMON PLOT THEME

surveillance_theme <- theme_minimal(
  base_size = 12
) +
  theme(
    plot.background =
      element_rect(
        fill = "#F6F4EF",
        colour = NA
      ),

    panel.background =
      element_rect(
        fill = "#F6F4EF",
        colour = NA
      ),

    panel.grid.minor =
      element_blank(),

    panel.grid.major.x =
      element_blank(),

    panel.grid.major.y =
      element_line(
        colour = "#DDDAD3",
        linewidth = 0.4
      ),

    axis.text =
      element_text(
        colour = "#65706C"
      ),

    axis.title =
      element_text(
        colour = "#65706C",
        size = 10
      ),

    strip.text =
      element_text(
        colour = "#26312D",
        face = "bold",
        size = 11
      ),

    strip.background =
      element_blank(),

    legend.position =
      "none",

    plot.title =
      element_text(
        colour = "#17201D",
        face = "bold",
        size = 15
      ),

    plot.subtitle =
      element_text(
        colour = "#65706C",
        size = 11
      )
  )

```


Overview
=====================================

Row {data-height=270}
-------------------------------------

```{r}

div(
  class = "hero",

  div(
    class = "eyebrow",
    "RESPIRATORY SURVEILLANCE · ENGLAND"
  ),

  tags$h1(
    "What's circulating this week?"
  ),

  div(
    class = "hero-copy",
    paste(
      "Weekly laboratory surveillance across six respiratory viruses,",
      "with statistical signal detection, short-term forecasting and",
      "NHS 111 syndromic surveillance."
    )
  ),

  div(
    class = "update-line",
    paste0(
      "Latest laboratory data · ",
      format(
        latest_date,
        "%d %B %Y"
      ),
      "  ·  UKHSA public surveillance data"
    )
  )
)

```


Row {data-height=155}
-------------------------------------

```{r}

tagList(

  div(
    class = "metric-grid",

    div(
      class = "metric-card",

      div(
        class = "metric-label",
        "Pathogens monitored"
      ),

      div(
        class = "metric-value",
        n_distinct(
          respiratory$disease
        )
      ),

      div(
        class = "metric-note",
        "Weekly PCR positivity"
      )
    ),


    div(
      class = "metric-card",

      div(
        class = "metric-label",
        "Current signals"
      ),

      div(
        class = "metric-value",
        current_signal_count
      ),

      div(
        class = "metric-note",
        "Above historical seasonal threshold"
      )
    ),


    div(
      class = "metric-card",

      div(
        class = "metric-label",
        "Surveillance history"
      ),

      div(
        class = "metric-value",
        n_distinct(
          respiratory$date
        )
      ),

      div(
        class = "metric-note",
        "Weeks of laboratory data"
      )
    ),


    div(
      class = "metric-card",

      div(
        class = "metric-label",
        "NHS 111 ARI calls"
      ),

      div(
        class = "metric-value",
        comma(
          latest_nhs111$weekly_calls
        )
      ),

      div(
        class = "metric-note",
        "Latest aligned complete week"
      )
    )
  )
)

```


Row {data-height=110}
-------------------------------------

```{r}

div(
  class = "section-intro",

  div(
    class = "section-label",
    "CURRENT ACTIVITY"
  ),

  tags$h2(
    "Six respiratory viruses, at a glance."
  ),

  tags$p(
    "Latest reported PCR positivity and statistical status relative to historical seasonal activity."
  )
)

```


Row {data-height=360}
-------------------------------------

```{r}

card_data <- signals |>
  mutate(
    disease =
      factor(
        disease,
        levels = pathogen_order
      )
  ) |>
  arrange(
    disease
  )


cards <- lapply(
  seq_len(
    nrow(card_data)
  ),
  function(i) {

    row <- card_data[i, ]

    disease_name <-
      as.character(
        row$disease
      )

    card_colour <-
      unname(
        pathogen_palette[
          disease_name
        ]
      )

    status_text <-
      if (
        is.na(
          row$signal
        )
      ) {

        "Baseline unavailable"

      } else if (
        row$signal
      ) {

        "Above expected"

      } else {

        "Expected range"

      }


    status_class <-
      if (
        is.na(
          row$signal
        )
      ) {

        "status status-unknown"

      } else if (
        row$signal
      ) {

        "status status-signal"

      } else {

        "status status-normal"

      }


    div(
      class = "pathogen-card",

      style =
        paste0(
          "border-top: 3px solid ",
          card_colour,
          ";"
        ),

      div(
        class = "pathogen-name",
        disease_name
      ),

      div(
        class = "pathogen-value",

        sprintf(
          "%.2f%%",
          row$positivity
        ),

        tags$span(
          class = "pathogen-unit",
          " PCR positivity"
        )
      ),

      tags$span(
        class = status_class,
        status_text
      )
    )
  }
)


div(
  class = "pathogen-grid",
  tagList(
    cards
  )
)

```


Row {data-height=460}
-------------------------------------

### Current activity

```{r}

latest_plot_data <- latest_data |>
  mutate(
    disease =
      factor(
        disease,
        levels = pathogen_order
      )
  )


activity_plot <- ggplot(
  latest_plot_data,
  aes(
    x = positivity,
    y = reorder(
      disease,
      positivity
    ),
    colour = disease
  )
) +

  geom_segment(
    aes(
      x = 0,
      xend = positivity,
      yend = reorder(
        disease,
        positivity
      )
    ),
    linewidth = 2.2,
    alpha = 0.35
  ) +

  geom_point(
    size = 4
  ) +

  geom_text(
    aes(
      label =
        sprintf(
          "%.2f%%",
          positivity
        )
    ),
    colour = "#26312D",
    hjust = -0.25,
    size = 3.6
  ) +

  scale_colour_manual(
    values = pathogen_palette
  ) +

  scale_x_continuous(
    labels =
      function(x) {
        paste0(
          x,
          "%"
        )
      },

    expand =
      expansion(
        mult = c(
          0,
          0.18
        )
      )
  ) +

  labs(
    x = "PCR positivity",
    y = NULL
  ) +

  surveillance_theme


ggplotly(
  activity_plot
)

```


Row {data-height=120}
-------------------------------------

```{r}

if (
  current_signal_count == 0
) {

  div(
    class = "note-box",

    tags$strong(
      "Current surveillance note. "
    ),

    "No pathogen currently exceeds its historical seasonal 95th percentile threshold."
  )

} else {

  div(
    class = "note-box signal-box",

    tags$strong(
      "Current surveillance note. "
    ),

    paste0(
      paste(
        current_signal_names,
        collapse = ", "
      ),
      ifelse(
        current_signal_count == 1,
        " is",
        " are"
      ),
      " currently above the historical seasonal threshold. ",
      "This is a statistical signal and is not equivalent to a confirmed outbreak."
    )
  )
}

```


Trends
=====================================

Row {data-height=125}
-------------------------------------

```{r}

div(
  class = "section-intro",

  div(
    class = "section-label",
    "SURVEILLANCE TRENDS"
  ),

  tags$h2(
    "How has activity changed over the last year?"
  ),

  tags$p(
    "Weekly PCR positivity for each respiratory pathogen over the most recent 52 weeks."
  )
)

```


Row {data-height=720}
-------------------------------------

### Recent pathogen activity

```{r}

recent_plot <- ggplot(
  recent_data,
  aes(
    x = date,
    y = positivity,
    colour = disease
  )
) +

  geom_line(
    linewidth = 0.85
  ) +

  facet_wrap(
    ~ disease,
    scales = "free_y",
    ncol = 2
  ) +

  scale_colour_manual(
    values = pathogen_palette
  ) +

  scale_x_date(
    date_labels = "%b",
    date_breaks = "2 months"
  ) +

  labs(
    x = NULL,
    y = "PCR positivity (%)"
  ) +

  surveillance_theme


ggplotly(
  recent_plot
)

```


Signals
=====================================

Row {data-height=125}
-------------------------------------

```{r}

div(
  class = "section-intro",

  div(
    class = "section-label",
    "STATISTICAL SURVEILLANCE"
  ),

  tags$h2(
    "Is anything behaving unusually?"
  ),

  tags$p(
    paste(
      "Observed activity is compared with the historical distribution",
      "for the corresponding point in the respiratory season."
    )
  )
)

```


Row {data-height=700}
-------------------------------------

### Observed activity and seasonal thresholds

```{r}

signal_plot <- ggplot(
  signal_recent,
  aes(
    x = date
  )
) +

  geom_line(
    aes(
      y = upper_threshold
    ),
    colour = "#A8ADA9",
    linewidth = 0.7,
    linetype = "dashed"
  ) +

  geom_line(
    aes(
      y = positivity,
      colour = disease
    ),
    linewidth = 0.9
  ) +

  geom_point(
    data =
      signal_recent |>
      filter(
        signal %in% TRUE
      ),

    aes(
      y = positivity
    ),

    colour = "#A94036",
    fill = "#F6F4EF",
    shape = 21,
    stroke = 1.2,
    size = 3
  ) +

  facet_wrap(
    ~ disease,
    scales = "free_y",
    ncol = 2
  ) +

  scale_colour_manual(
    values = pathogen_palette
  ) +

  scale_x_date(
    date_labels = "%b",
    date_breaks = "2 months"
  ) +

  labs(
    x = NULL,
    y = "PCR positivity (%)",
    caption =
      "Dashed line = historical seasonal 95th percentile · highlighted points = statistical signals"
  ) +

  surveillance_theme


ggplotly(
  signal_plot
)

```


Row {data-height=300}
-------------------------------------

### Latest signal status

```{r}

signal_table <- signals |>
  transmute(
    Pathogen =
      disease,

    Week =
      date,

    `Observed (%)` =
      round(
        positivity,
        2
      ),

    `Expected (%)` =
      round(
        expected,
        2
      ),

    `Threshold (%)` =
      round(
        upper_threshold,
        2
      ),

    Status =
      case_when(
        is.na(signal) ~
          "Baseline unavailable",

        signal ~
          "Above expected",

        TRUE ~
          "Expected range"
      )
  )


datatable(
  signal_table,
  rownames = FALSE,
  options = list(
    dom = "t",
    pageLength = 6,
    ordering = FALSE
  )
)

```


Row {data-height=150}
-------------------------------------

<div class="note-box">

<strong>How to read this.</strong>

A signal occurs when observed positivity exceeds the historical 95th
percentile for the corresponding week of the respiratory season.
Signals identify unusual statistical activity for further investigation;
they do not confirm that an outbreak is occurring.

</div>


Forecast
=====================================

Row {data-height=125}
-------------------------------------

```{r}

div(
  class = "section-intro",

  div(
    class = "section-label",
    "SHORT-TERM OUTLOOK"
  ),

  tags$h2(
    "What might happen over the next four weeks?"
  ),

  tags$p(
    paste(
      "Model-based projections generated from surveillance available",
      "up to the latest reporting week."
    )
  )
)

```


Row {data-height=750}
-------------------------------------

### Observed activity and ARIMA projections

```{r}

forecast_plot <- ggplot() +

  geom_line(
    data =
      forecast_history,

    aes(
      x = date,
      y = positivity,
      colour = disease
    ),

    linewidth = 0.85
  ) +

  geom_ribbon(
    data =
      current_forecasts,

    aes(
      x = date,
      ymin = lower_95,
      ymax = upper_95
    ),

    fill = "#8F9995",
    alpha = 0.10
  ) +

  geom_ribbon(
    data =
      current_forecasts,

    aes(
      x = date,
      ymin = lower_80,
      ymax = upper_80
    ),

    fill = "#77827D",
    alpha = 0.16
  ) +

  geom_line(
    data =
      current_forecasts,

    aes(
      x = date,
      y = forecast,
      colour = disease
    ),

    linetype = "dashed",
    linewidth = 0.95
  ) +

  facet_wrap(
    ~ disease,
    scales = "free_y",
    ncol = 2
  ) +

  scale_colour_manual(
    values = pathogen_palette
  ) +

  scale_x_date(
    date_labels = "%b",
    date_breaks = "2 months"
  ) +

  labs(
    x = NULL,
    y = "PCR positivity (%)",
    caption =
      "Solid line = observed · dashed line = forecast · shading = 80% and 95% prediction intervals"
  ) +

  surveillance_theme


ggplotly(
  forecast_plot
)

```


Row {data-height=330}
-------------------------------------

### Forecast values

```{r}

forecast_table <- current_forecasts |>
  transmute(
    Pathogen =
      disease,

    Week =
      date,

    Horizon =
      paste0(
        horizon,
        ifelse(
          horizon == 1,
          " week",
          " weeks"
        )
      ),

    Forecast =
      round(
        forecast,
        2
      ),

    `80% lower` =
      round(
        lower_80,
        2
      ),

    `80% upper` =
      round(
        upper_80,
        2
      )
  )


datatable(
  forecast_table,
  rownames = FALSE,
  options = list(
    pageLength = 12
  )
)

```


Row {data-height=155}
-------------------------------------

<div class="note-box">

<strong>Forecast interpretation.</strong>

ARIMA is used here as a simple current operational forecasting model.
In rolling-origin evaluation during 2025/26, ARIMA produced the lowest
MAE in 19 of 24 pathogen-horizon comparisons. Performance nevertheless
varied by pathogen and horizon, and seasonal-naïve forecasting
outperformed ARIMA for adenovirus at several longer horizons.

</div>


NHS 111
=====================================

Row {data-height=125}
-------------------------------------

```{r}

div(
  class = "section-intro",

  div(
    class = "section-label",
    "SYNDROMIC SURVEILLANCE"
  ),

  tags$h2(
    "What are NHS 111 respiratory calls showing?"
  ),

  tags$p(
    paste(
      "Acute respiratory infection calls provide a broader measure",
      "of respiratory healthcare activity alongside laboratory surveillance."
    )
  )
)

```


Row {data-height=155}
-------------------------------------

```{r}

tagList(

  div(
    class = "metric-grid",

    div(
      class = "metric-card",

      div(
        class = "metric-label",
        "Latest weekly calls"
      ),

      div(
        class = "metric-value",
        comma(
          latest_nhs111$weekly_calls
        )
      ),

      div(
        class = "metric-note",
        format(
          latest_nhs111$week_start,
          "%d %B %Y"
        )
      )
    ),


    div(
      class = "metric-card",

      div(
        class = "metric-label",
        "Week-on-week change"
      ),

      div(
        class = "metric-value",

        paste0(
          ifelse(
            latest_nhs111_change$percent_change > 0,
            "+",
            ""
          ),

          round(
            latest_nhs111_change$percent_change,
            1
          ),

          "%"
        )
      ),

      div(
        class = "metric-note",
        "Compared with previous week"
      )
    )
  )
)

```


Row {data-height=470}
-------------------------------------

### NHS 111 acute respiratory infection activity

```{r}

nhs111_plot <- syndromic |>
  filter(
    week_start >=
      max(
        week_start
      ) -
      weeks(
        52
      )
  ) |>
  ggplot(
    aes(
      x = week_start,
      y = weekly_calls
    )
  ) +

  geom_line(
    colour = "#327565",
    linewidth = 1
  ) +

  geom_point(
    data =
      latest_nhs111,

    aes(
      x = week_start,
      y = weekly_calls
    ),

    colour = "#327565",
    size = 3
  ) +

  scale_y_continuous(
    labels = comma
  ) +

  scale_x_date(
    date_labels = "%b",
    date_breaks = "2 months"
  ) +

  labs(
    x = NULL,
    y = "Weekly NHS 111 calls"
  ) +

  surveillance_theme


ggplotly(
  nhs111_plot
)

```


Row {data-height=115}
-------------------------------------

```{r}

div(
  class = "section-intro",

  div(
    class = "section-label",
    "LEAD–LAG ANALYSIS"
  ),

  tags$h2(
    "Does syndromic activity give earlier warning?"
  )
)

```


Row {data-height=430}
-------------------------------------

### Week-to-week changes

```{r}

lead_lag_plot <- ggplot(
  lead_lag,
  aes(
    x = lag_weeks,
    y = correlation
  )
) +

  geom_hline(
    yintercept = 0,
    colour = "#BFC2BE",
    linewidth = 0.5
  ) +

  geom_vline(
    xintercept = 0,
    colour = "#7B8581",
    linetype = "dotted",
    linewidth = 0.7
  ) +

  geom_line(
    colour = "#327565",
    linewidth = 1
  ) +

  geom_point(
    colour = "#327565",
    size = 3
  ) +

  scale_x_continuous(
    breaks = -6:6
  ) +

  labs(
    x = "Lag (weeks)",
    y = "Correlation"
  ) +

  surveillance_theme


ggplotly(
  lead_lag_plot
)

```


Row {data-height=190}
-------------------------------------

<div class="note-box">

<strong>What did we find?</strong>

Across 134 overlapping weeks, the strongest association between
week-to-week changes in NHS 111 acute respiratory infection calls and
influenza PCR positivity occurred contemporaneously at <strong>lag 0
(r = 0.498)</strong>.

Season-specific results differed: the peak occurred at approximately
+1 week in 2024/25 and -1 week in 2025/26. The available data therefore
do not demonstrate a consistent syndromic lead time.

</div>


Methods
=====================================

Row {data-height=120}
-------------------------------------

```{r}

div(
  class = "section-intro",

  div(
    class = "section-label",
    "METHODOLOGY"
  ),

  tags$h2(
    "How this surveillance system works."
  ),

  tags$p(
    "A reproducible R workflow built using publicly available UKHSA surveillance data."
  )
)

```


Row
-------------------------------------

<div class="method-card">

<div class="section-label">DATA</div>

### Laboratory surveillance

Weekly PCR positivity is monitored for influenza, RSV, rhinovirus,
human metapneumovirus, adenovirus and parainfluenza in England.

The cleaned laboratory dataset contains 481 weekly observations per
pathogen from July 2017 onwards.

</div>


<div class="method-card">

<div class="section-label">SIGNAL DETECTION</div>

### Identifying unusual activity

For each pathogen and week of the respiratory season, historical
observations are used to calculate a median expected value and empirical
95th percentile threshold.

Activity above this threshold is classified as a statistical signal.

The method was subsequently evaluated using leave-one-season-out
retrospective validation to avoid evaluating historical observations
against a baseline containing those same observations.

</div>


<div class="method-card">

<div class="section-label">FORECASTING</div>

### Looking 1–4 weeks ahead

Seasonal naïve, ARIMA and STL-ETS forecasting approaches were compared
using rolling weekly forecast origins.

During the 2025/26 evaluation period, ARIMA achieved the lowest MAE in
19 of 24 pathogen-horizon comparisons.

This dashboard uses ARIMA for current projections as a practical
simplification rather than assuming that one model is optimal for every
pathogen.

</div>


<div class="method-card">

<div class="section-label">SYNDROMIC SURVEILLANCE</div>

### NHS 111 acute respiratory infection calls

Daily NHS 111 acute respiratory infection calls were aggregated to
complete epidemiological weeks and aligned with influenza laboratory
surveillance.

NHS 111 activity represents broad respiratory healthcare-seeking
behaviour and is not specific to influenza.

</div>


<div class="method-card">

<div class="section-label">LIMITATIONS</div>

### How the dashboard should be interpreted

This is an exploratory epidemiological surveillance and data-science
project rather than an operational public-health alerting system.

Statistical signals indicate unusual observations rather than confirmed
outbreaks. Forecast performance may vary between seasons, PCR positivity
can be affected by testing behaviour and surveillance practices, and
only two complete respiratory seasons were available for the
season-specific NHS 111 lead-lag analysis.

</div>


<div class="method-card">

<div class="section-label">ABOUT THIS PROJECT</div>

### Reproducible public-health surveillance in R

Data are derived from public UK Health Security Agency surveillance
sources. This dashboard is an independent portfolio project and is not
an official UKHSA product or public-health advisory.

</div>