---
title: "England Respiratory Surveillance"
subtitle: "Multi-pathogen surveillance, statistical signals and short-term forecasting"
author: "Chandhini Suresh"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    theme: cosmo
---

```{r setup, include=FALSE}

# 1. PACKAGES

library(flexdashboard)
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(ggplot2)
library(plotly)
library(forecast)
library(DT)
library(scales)

```


```{r data, include=FALSE}

# 2. LOAD RESPIRATORY DATA

respiratory <- read_csv(
  "data/processed/respiratory_surveillance_long.csv",
  show_col_types = FALSE
) |>
  mutate(
    date = as.Date(date)
  )


# 3. CREATE RESPIRATORY SEASON VARIABLES

respiratory_signals <- respiratory |>
  mutate(
    season_start_year =
      if_else(
        month(date) >= 7,
        year(date),
        year(date) - 1
      ),

    season =
      paste0(
        season_start_year,
        "/",
        substr(
          season_start_year + 1,
          3,
          4
        )
      ),

    season_start =
      as.Date(
        paste0(
          season_start_year,
          "-07-01"
        )
      ),

    week_of_season =
      floor(
        as.numeric(
          date - season_start
        ) / 7
      ) + 1
  )


# 4. IDENTIFY CURRENT RESPIRATORY SEASON

current_season <- respiratory_signals |>
  filter(
    date == max(date)
  ) |>
  distinct(
    season
  ) |>
  pull(
    season
  )


# 5. BUILD HISTORICAL SIGNAL BASELINE

historical_baseline <- respiratory_signals |>
  filter(
    season != current_season
  ) |>
  group_by(
    disease,
    week_of_season
  ) |>
  summarise(
    expected =
      median(
        positivity,
        na.rm = TRUE
      ),

    upper_threshold =
      quantile(
        positivity,
        probs = 0.95,
        na.rm = TRUE
      ),

    .groups = "drop"
  )


# 6. APPLY SIGNAL DETECTOR

signals_all <- respiratory_signals |>
  left_join(
    historical_baseline,
    by = c(
      "disease",
      "week_of_season"
    )
  ) |>
  mutate(
    signal =
      positivity > upper_threshold
  )


# 7. LATEST SIGNAL STATUS

signals <- signals_all |>
  group_by(
    disease
  ) |>
  slice_max(
    date,
    n = 1,
    with_ties = FALSE
  ) |>
  ungroup()


# 8. LATEST LABORATORY OBSERVATION

latest_date <- max(
  respiratory$date,
  na.rm = TRUE
)


latest_data <- respiratory |>
  filter(
    date == latest_date
  )


# 9. RECENT LABORATORY DATA

recent_start <-
  latest_date - weeks(52)


recent_data <- respiratory |>
  filter(
    date >= recent_start
  )


# 10. LOAD SYNDROMIC DATA

syndromic <- read_csv(
  "data/processed/syndromic_influenza_aligned.csv",
  show_col_types = FALSE
) |>
  mutate(
    week_start =
      as.Date(week_start)
  )


# 11. LATEST NHS 111 OBSERVATION

latest_nhs111 <- syndromic |>
  slice_max(
    week_start,
    n = 1,
    with_ties = FALSE
  )


# 12. NHS 111 WEEKLY CHANGE

nhs111_recent <- syndromic |>
  arrange(
    week_start
  ) |>
  mutate(
    weekly_change =
      weekly_calls -
      lag(weekly_calls),

    percent_change =
      100 *
      weekly_change /
      lag(weekly_calls)
  )


latest_nhs111_change <- nhs111_recent |>
  slice_max(
    week_start,
    n = 1,
    with_ties = FALSE
  )


# 13. CURRENT FORECAST FUNCTION

forecast_current <- function(data) {

  data <- data |>
    arrange(
      date
    )

  disease_name <-
    unique(
      data$disease
    )

  y <- ts(
    data$positivity,
    frequency = 52
  )

  model <- auto.arima(
    y,
    seasonal = TRUE,
    stepwise = TRUE,
    approximation = TRUE
  )

  fc <- forecast(
    model,
    h = 4,
    level = c(
      80,
      95
    )
  )

  last_date <-
    max(
      data$date
    )

  forecast_dates <- seq(
    last_date + weeks(1),
    by = "1 week",
    length.out = 4
  )

  tibble(
    disease =
      disease_name,

    date =
      forecast_dates,

    horizon =
      1:4,

    forecast =
      as.numeric(
        fc$mean
      ),

    lower_80 =
      as.numeric(
        fc$lower[, "80%"]
      ),

    upper_80 =
      as.numeric(
        fc$upper[, "80%"]
      ),

    lower_95 =
      as.numeric(
        fc$lower[, "95%"]
      ),

    upper_95 =
      as.numeric(
        fc$upper[, "95%"]
      )
  )
}


# 14. GENERATE CURRENT FORECASTS

current_forecasts <- respiratory |>
  group_by(
    disease
  ) |>
  group_split() |>
  lapply(
    forecast_current
  ) |>
  bind_rows()


# 15. LOAD LEAD-LAG RESULTS

lead_lag <- read_csv(
  "outputs/tables/change_lag_correlations.csv",
  show_col_types = FALSE
)

```


Overview
=====================================

Row
-------------------------------------

### Latest surveillance week

```{r}

valueBox(
  format(
    latest_date,
    "%d %B %Y"
  ),
  caption =
    "Latest laboratory surveillance week",
  icon =
    "fa-calendar"
)

```


### Respiratory pathogens

```{r}

valueBox(
  n_distinct(
    respiratory$disease
  ),
  caption =
    "Pathogens monitored",
  icon =
    "fa-chart-line"
)

```


### Current statistical signals

```{r}

current_signal_count <- signals |>
  filter(
    signal == TRUE
  ) |>
  nrow()


valueBox(
  current_signal_count,
  caption =
    "Current statistical signals",
  icon =
    "fa-exclamation-triangle"
)

```


### Latest NHS 111 ARI calls

```{r}

valueBox(
  comma(
    latest_nhs111$weekly_calls
  ),
  caption =
    "Latest complete week",
  icon =
    "fa-phone"
)

```


Row
-------------------------------------

### Current respiratory virus activity

```{r}

latest_plot <- ggplot(
  latest_data,
  aes(
    x = reorder(
      disease,
      positivity
    ),
    y = positivity
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "PCR positivity (%)"
  ) +
  theme_minimal()


ggplotly(
  latest_plot
)

```


### Latest surveillance table

```{r}

latest_table <- latest_data |>
  select(
    disease,
    positivity
  ) |>
  arrange(
    desc(positivity)
  ) |>
  mutate(
    positivity =
      round(
        positivity,
        2
      )
  )


datatable(
  latest_table,
  rownames = FALSE,
  options = list(
    dom = "t",
    pageLength = 6
  )
)

```


Recent Trends
=====================================

Row
-------------------------------------

### Last 12 months

```{r}

recent_plot <- ggplot(
  recent_data,
  aes(
    x = date,
    y = positivity
  )
) +
  geom_line(
    linewidth = 0.7
  ) +
  facet_wrap(
    ~ disease,
    scales = "free_y",
    ncol = 2
  ) +
  labs(
    x = NULL,
    y = "PCR positivity (%)"
  ) +
  theme_minimal()


ggplotly(
  recent_plot
)

```


Signals
=====================================

Row
-------------------------------------

### Current statistical signal status

```{r}

signal_table <- signals |>
  select(
    disease,
    date,
    positivity,
    expected,
    upper_threshold,
    signal
  ) |>
  mutate(
    positivity =
      round(
        positivity,
        2
      ),

    expected =
      round(
        expected,
        2
      ),

    upper_threshold =
      round(
        upper_threshold,
        2
      )
  ) |>
  arrange(
    desc(signal),
    disease
  )


datatable(
  signal_table,
  rownames = FALSE,
  options = list(
    dom = "t",
    pageLength = 6
  )
)

```


Row
-------------------------------------

### Signal interpretation

A statistical signal occurs when observed positivity exceeds the
historical season-specific 95th percentile threshold.

These are provisional statistical signals and should not be
interpreted as confirmation of an outbreak.

The signal-detection approach was retrospectively evaluated using
leave-one-season-out validation.


Forecasts
=====================================

Row
-------------------------------------

### Current 1–4 week forecasts

```{r}

forecast_history <- respiratory |>
  filter(
    date >=
      latest_date - weeks(26)
  )


forecast_plot <- ggplot() +

  geom_line(
    data =
      forecast_history,

    aes(
      x = date,
      y = positivity
    ),

    linewidth = 0.7
  ) +

  geom_ribbon(
    data =
      current_forecasts,

    aes(
      x = date,
      ymin = lower_95,
      ymax = upper_95
    ),

    alpha = 0.15
  ) +

  geom_ribbon(
    data =
      current_forecasts,

    aes(
      x = date,
      ymin = lower_80,
      ymax = upper_80
    ),

    alpha = 0.25
  ) +

  geom_line(
    data =
      current_forecasts,

    aes(
      x = date,
      y = forecast
    ),

    linetype = "dashed",
    linewidth = 0.8
  ) +

  facet_wrap(
    ~ disease,
    scales = "free_y",
    ncol = 2
  ) +

  labs(
    x = NULL,

    y =
      "PCR positivity (%)",

    caption =
      "Dashed line: ARIMA forecast | Shading: 80% and 95% prediction intervals"
  ) +

  theme_minimal()


ggplotly(
  forecast_plot
)

```


Row
-------------------------------------

### Forecast values

```{r}

forecast_table <- current_forecasts |>
  mutate(
    across(
      c(
        forecast,
        lower_80,
        upper_80,
        lower_95,
        upper_95
      ),
      ~ round(
        .x,
        2
      )
    )
  )


datatable(
  forecast_table,
  rownames = FALSE,
  options = list(
    pageLength = 12
  )
)

```


Syndromic Surveillance
=====================================

Row
-------------------------------------

### NHS 111 acute respiratory infection activity

```{r}

nhs111_plot <- syndromic |>
  filter(
    week_start >=
      max(week_start) - weeks(52)
  ) |>
  ggplot(
    aes(
      x = week_start,
      y = weekly_calls
    )
  ) +
  geom_line(
    linewidth = 0.8
  ) +
  labs(
    x = NULL,
    y = "Weekly NHS 111 calls"
  ) +
  theme_minimal()


ggplotly(
  nhs111_plot
)

```


### Latest weekly change

```{r}

valueBox(
  paste0(
    round(
      latest_nhs111_change$percent_change,
      1
    ),
    "%"
  ),
  caption =
    "Change in NHS 111 ARI calls from previous week",
  icon =
    "fa-chart-line"
)

```


Lead-Lag Analysis
=====================================

Row
-------------------------------------

### Week-to-week changes

```{r}

lead_lag_plot <- ggplot(
  lead_lag,
  aes(
    x = lag_weeks,
    y = correlation
  )
) +

  geom_hline(
    yintercept = 0,
    linetype = "dashed"
  ) +

  geom_vline(
    xintercept = 0,
    linetype = "dotted"
  ) +

  geom_line(
    linewidth = 0.8
  ) +

  geom_point(
    size = 2.5
  ) +

  scale_x_continuous(
    breaks = -6:6
  ) +

  labs(
    x = "Lag (weeks)",
    y = "Pearson correlation"
  ) +

  theme_minimal()


ggplotly(
  lead_lag_plot
)

```


### Interpretation

Across the available overlapping surveillance period, the strongest
association between week-to-week changes in NHS 111 acute respiratory
infection calls and influenza PCR positivity occurred at lag 0.

This indicates that the strongest observed relationship was
contemporaneous.

Season-specific analysis identified a peak at approximately +1 week
in 2024/25 and -1 week in 2025/26.

The available data therefore do not demonstrate a consistent syndromic
lead time.


Methods & Limitations
=====================================

### Data

Laboratory and syndromic surveillance data were obtained from the
UK Health Security Agency dashboard API.

The laboratory surveillance dataset contains weekly PCR positivity for
six respiratory pathogens:

- Influenza
- Respiratory syncytial virus
- Human metapneumovirus
- Rhinovirus
- Adenovirus
- Parainfluenza


### Statistical signal detection

Expected activity was calculated from historical observations for the
corresponding week of the respiratory season.

Observations above the empirical historical 95th percentile were
classified as statistical signals.

This detector is exploratory and was retrospectively evaluated using
leave-one-season-out validation. Signals should not be interpreted as
confirmed outbreaks.


### Forecasting

Forecasts are generated for 1–4 weeks ahead.

Three forecasting approaches were evaluated previously using
rolling-origin validation:

- Seasonal naïve
- ARIMA
- STL-ETS

During the 2025/26 evaluation, ARIMA produced the lowest MAE in
19 of 24 pathogen-horizon comparisons.

Performance varied by pathogen and forecast horizon. In particular,
seasonal naïve forecasting outperformed ARIMA for adenovirus at several
longer horizons.

The dashboard uses ARIMA as a simple current operational forecasting
model rather than implying that it is universally optimal.


### Syndromic surveillance

NHS 111 acute respiratory infection calls measure syndromic healthcare
activity rather than laboratory-confirmed influenza.

The overlapping NHS 111 and influenza dataset contains 134 weeks
from February 2024 onwards.

Only two complete respiratory seasons were available for
season-specific lead-lag comparison.


### Interpretation

This dashboard is an exploratory epidemiological surveillance project.

Statistical signals indicate unusual observations rather than confirmed
outbreaks.

Forecasts contain uncertainty and should be interpreted alongside
observed surveillance data and wider epidemiological evidence.

Lead-lag correlations describe temporal association and do not establish
causality or a consistent predictive lead.